---
title: "nflplotR & the gt Package"
description: "How nflplotR works as gt extension"
author: "Sebastian Carl"
opengraph:
  image: 
    src: "https://raw.githubusercontent.com/nflverse/nflplotR/main/man/figures/social_preview_gt.png"
  twitter:
    card: summary_large_image
    creator: "@mrcaseb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nflplotR & the gt Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 600
)
options(nflreadr.cache_warning = FALSE)
options(warn = -1)
```

The R package [{gt}](https://gt.rstudio.com/) is becoming increasingly popular for creating aesthetically pleasing tables. nflplotR supports rendering of team logos, team wordmarks, and player headshots in gt tables similar to ggplot2. This article will provide some typical examples. 

## Team Logos & Wordmarks

The functions `gt_nfl_logos()` and `gt_nfl_wordmarks()` come with a powerful `locations` argument that allows usage of gt selection helpers. We will create an example dataframe to show how this all works.

```{r}
df <- data.frame(
  row_group_column = c("AFC", "NFC", "AFC", "NFC"),
  row_name_column = c("LAC", "SEA"),
  column_a = 11:12,
  column_b = c("KC", "LA")
) 
```

Our example dataframe in a gt table without any formatting.

```{r}
gt::gt(df)
```

The column `row_group_column` is intended to serve as row group variable so let's apply this.

```{r}
gt::gt(df, groupname_col = "row_group_column")
```

We also would like to render images in the stub, i.e. the rownames so we tell gt about the `row_name_column`. 

```{r}
example_table <- gt::gt(
  df, 
  groupname_col = "row_group_column", 
  rowname_col = "row_name_column"
) |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
example_table
```

This is our final table. We have valid NFL abbreviations in the cell body, in row group labels and in the stub. We can now use nflplotR to render images instead of those abbreviations.

### Cell Body

To render images in the cell body, i.e. the rows of the table, we can either use the `columns` argument or the appropriate `locations` helper.

```{r}
example_table |> 
  nflplotR::gt_nfl_logos(columns = "column_b")
```

Please note, that the locations helper will allow you to selectively apply the function to a set of rows

```{r}
example_table |> 
  nflplotR::gt_nfl_logos(locations = gt::cells_body(rows = gt::starts_with("LAC")))
```


### Row Group Label

Rendering images outside of the cell body will always require the appropriate call to the `locations` argument. The `columns` argument cannot handle anything outside cell bodies.

```{r}
example_table |> 
  nflplotR::gt_nfl_logos(locations = gt::cells_row_groups())
```

### Stub

Now we would like to convert rownames to images.

```{r}
example_table |> 
  nflplotR::gt_nfl_wordmarks(locations = gt::cells_stub())
```

### Column Spanners

We use another table to demonstrate how to convert team names to logos and wordmarks in column spanners.

Here only logos

```{r}
df <- data.frame(a = 3, b = 4, c = 5, d = 6)

df |> 
  gt::gt() |> 
  gt::tab_spanner("KC", c(a, b)) |> 
  gt::tab_spanner("LAC", c(c, d)) |> 
  nflplotR::gt_nfl_logos(locations = gt::cells_column_spanners())
```

And now mix logo and wordmark

```{r}
df <- data.frame(a = 3, b = 4, c = 5, d = 6)

df |> 
  gt::gt() |> 
  gt::tab_spanner("KC", c(a, b)) |> 
  gt::tab_spanner("LAC", c(c, d)) |> 
  nflplotR::gt_nfl_logos(locations = gt::cells_column_spanners("KC")) |> 
  nflplotR::gt_nfl_wordmarks(locations = gt::cells_column_spanners("LAC"))
```

### Combine all together

The `locations` argument allows multiple locations in one call by wrapping them in a list.

```{r}
example_table |> 
  nflplotR::gt_nfl_wordmarks(locations = gt::cells_stub()) |> 
  nflplotR::gt_nfl_logos(
    locations = list(
      gt::cells_body(), gt::cells_row_groups()
    )
  )
```

### How about Column Labels?

Well...it's complicated, because [gt behaves inconsistent in my opinion](https://github.com/rstudio/gt/issues/1433).

The actually correct way would be a call to `nflplotR::gt_nfl_logos` or `nflplotR::gt_nfl_wordmarks` with the `locations` argument set to `gt::cells_column_labels()`. Currently, this wouldn't render any images in column labels as discussed in the above linked issue. 

However, as a convenient workaround, nflplotR supports logos and wordmarks in column labels through `gt_nfl_cols_label()`. 

LOGOS:

```{r}
teams <- nflplotR::valid_team_names() |> head(6)
df <- cbind(1, 2, 3, 4, 5, 6) |> 
  as.data.frame() |> 
  rlang::set_names(teams)

gt::gt(df) |> 
  nflplotR::gt_nfl_cols_label() |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
```

WORDMARKS (note how non matches remain unchanged):

```{r}
gt::gt(df) |> 
  nflplotR::gt_nfl_cols_label(type = "wordmark") |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
```

HEADSHOTS:

```{r}
headshot_df <- data.frame(
  "00-0036355" = 1, 
  "00-0033873" = 2,
  check.names = FALSE
)
gt::gt(headshot_df) |> 
  nflplotR::gt_nfl_cols_label(type = "headshot") |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
```

### Logos and Wordmarks Rendered by nflplotR

This example creates a table that renders all team logos and wordmarks. We split the table into 2 x 16 rows to avoid an overly long table and convert all variables starting with "logo" to logos and all variables starting with "wordmark" to wordmarks. 

```{r}
teams <- nflplotR::valid_team_names()
# remove conference logos from this example
teams <- teams[!teams %in% c("AFC", "NFC", "NFL")]
# create dataframe with all 32 team names
df <- data.frame(
  team_a = head(teams, 16),
  logo_a = head(teams, 16),
  wordmark_a = head(teams, 16),
  team_b = tail(teams, 16),
  logo_b = tail(teams, 16),
  wordmark_b = tail(teams, 16)
)
# create gt table and translate team names to logo/wordmark images
df |>
  gt::gt() |>
  nflplotR::gt_nfl_logos(columns = gt::starts_with("logo")) |>
  nflplotR::gt_nfl_wordmarks(columns = gt::starts_with("wordmark")) |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
```

## Player Headshots

All of the above applies to `gt_nfl_headshots()` as well. All you need is a gsis ID. 

```{r}
df <- data.frame(
  A = c("00-0036355", "00-0033873"),
  B = c("00-0033077", "00-0035228")
)

df |>
  gt::gt() |> 
  nflplotR::gt_nfl_headshots(columns = gt::everything(), height = 50) |> 
  # align the complete table left
  gt::tab_options(
    table.align = "left"
  )
```
